---
title: "Effective rank and weak identifiability in ff() terms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effective rank and weak identifiability in ff() terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A function-on-function term `ff(X, xind = s)` estimates a coefficient surface
$\beta(t,s)$, but the data only inform $\beta$ in directions along $s$ that the
observed curves $X_i(s)$ actually span. `ff()` measures that span by the
*effective rank* of $\mathrm{Cov}(X(s))$: the number of eigenvalues needed to
account for 99.5% of the variance in `X`. With `check.ident = TRUE` (the
default) `ff()` warns when this effective rank falls below `1.5 * k_s`, where
`k_s` is the marginal basis dimension along `s`.

```{r, message = FALSE}
library(refund)
set.seed(1)
s <- seq(0, 1, length.out = 40)
# 30 curves built from only 5 Fourier components: effective rank ~ 5
scores <- matrix(rnorm(30 * 5), 30, 5)
basis <- sapply(1:5, function(j) sin(j * pi * s))
X <- scores %*% t(basis)

term <- try(ff(X, xind = s, splinepars = list(bs = "ps", k = c(8, 5))))
```

## Why 1.5, and what the warning means

Identifiability here is not a yes/no property. The stricter of the two
classical conditions, effective rank $< k_s$, means the surface is pinned down
by the penalty alone. But the problem starts well before that point: as $k_s$
approaches the effective rank, a growing part of a rough true surface lies
outside the span of the observed curves, and the fit returns the penalty's
answer for that component instead of an estimate.

This is a *shared bias floor*. It is not specific to one estimator or one kind
of standard error, so it degrades point estimates and interval coverage
together, and — the reason for the warning — it produces no fitting failure and
no visible symptom. In the simulation studies behind `pffr()`'s cluster-robust
intervals, a margin of about a factor 1.5 was the smallest at which this
contamination was negligible; below it, several percent of the true surface's
energy was unreachable by any method.

## What to do about it

* **Lower `k_s`**, via `splinepars = list(bs = "ps", k = c(k_s, k_t))`. This is
  usually the right response: a basis the curves cannot support is not buying
  flexibility, only penalty-driven fill-in.
* **Use more, or more varied, curves.** The effective rank can never exceed
  `min(nrow(X), ncol(X))`, so with few curves the warning may be impossible to
  satisfy at the requested `k_s` — the design, not the code, is the binding
  constraint.
* **Switch to `ffpc()`**, which parameterizes the effect in the leading
  functional principal components of `X` and is built for the genuinely
  low-rank case.
* If none of these is available, the fit remains usable — but conclusions about
  $\beta(t,s)$, including the width and coverage of its confidence bands,
  inherit the caveat. Report it.

The check can be switched off with `check.ident = FALSE`.

See `?ff` and Scheipl, F., Greven, S. (2016). Identifiability in penalized
function-on-function regression models. *Electronic Journal of Statistics*
**10**(1), 495–526.
