---
title: "Pointwise Confidence Intervals for Penalized Function-on-Function Regression: A Simulation Study"
author:
  - name: Fabian Scheipl
    affiliations:
      - name: LMU München
        department: Department of Statistics
        city: Munich
        country: Germany
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
bibliography: pffr-ci-refs.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("_pffr-ci-helpers.R")
```

## Introduction {#sec-intro}

<!-- TODO: Motivate the practical importance of valid pointwise CIs for pffr().
     Key points: functional regression widely used in biomedical/environmental
     applications; inference quality under-studied compared to point estimation;
     default Bayesian CIs from mgcv may break under within-curve dependence. -->

Penalized functional regression models, as implemented by `pffr()` in the
`refund` package [@ScheiplStaicuGreven2015; @Rpkg_refund], are widely used for
analyzing functional data. These models leverage the penalized spline framework
of `mgcv` [@Wood2017] to estimate smooth functional effects, but the default
pointwise confidence intervals (CIs) rely on Bayesian posterior covariance
matrices that assume correctly specified error structures.

When within-curve dependence is present --- as is typical in longitudinal and
functional data settings --- these default CIs can severely undercover.
Sandwich-based covariance estimators offer a robust alternative
[@LiangZeger1986; @ZeileisKoellGraham2020], and bias-corrected variants such as
the CR2 estimator [@BellMcCaffrey2002; @PustejovskyTipton2018] can further
improve finite-sample performance.

This report presents two simulation studies evaluating pointwise CI methods for
`pffr()`, following the ADEMP framework for simulation study reporting
[@MorrisWhiteCrowther2019]. Study 1 examines non-Gaussian families (Poisson,
Binomial) under varying correlation structures. Study 2 investigates how
response grid resolution interacts with CI quality, revealing a previously
undocumented collapse of default standard errors on fine grids under correlation.


## Methods {#sec-methods}

### The pffr model {#sec-pffr}

<!-- TODO: Brief mathematical description of the pffr model class.
     Y_i(t) = eta_i(t) + eps_i(t), with eta combining intercept, ff, linear,
     smooth, and concurrent terms. Reference Wood 2017 for the penalized
     likelihood framework. -->

The `pffr()` function fits penalized function-on-function additive models of the
form
$$
Y_i(t) = \beta_0(t) + \int X_i(s)\beta(s,t)\,ds + z_i \beta_{\text{lin}}(t) + f(x_i, t) + X_i(t)\beta_{\text{conc}}(t) + \varepsilon_i(t),
$$
where $i = 1,\ldots,G$ indexes curves (clusters), $t$ is the functional domain,
$\beta(s,t)$ is a bivariate coefficient surface, and $\varepsilon_i(t)$ is the
error process. The model is fit as a large penalized generalized additive model
via `mgcv::gam()` [@Wood2017], with tensor product spline bases and appropriate
smoothing penalties.

### Sandwich covariance estimators {#sec-sandwich}

After basis expansion, `pffr()` stacks all $N = \sum_{i=1}^{G} m_i$ scalar
response evaluations (curve $i$ observed at $m_i$ grid points) into a single
penalized GAM with design matrix $X$ ($N \times p$), coefficient vector
$\hat\theta$ collecting the spline coefficients of every model term, and a
block-diagonal smoothing penalty $S = \sum_j \lambda_j S_j$. `mgcv` returns two
covariance matrices for $\hat\theta$: the **Bayesian posterior** covariance
$$
V_p = (X^\top W X + S)^{-1},
$$
with $W$ the (working) weight matrix, and the **frequentist** covariance
$V_e = V_p (X^\top W X) V_p$. Default `pffr()` CIs use $V_p$, which delivers good
*across-the-function* frequentist coverage under a correctly specified error
model because the term $V_p - V_e \succeq 0$ absorbs the smoothing-penalty
(prior) contribution to the sampling error [@MarraWood2012]. This guarantee,
however, presumes independent, correctly weighted residuals; it says nothing
about within-curve dependence, which is the norm in functional data.

All three robust estimators replace the model-based meat of the covariance while
retaining the *penalized* bread $V_p$ and the Bayesian correction term. They
share the form
$$
\widehat{V} \;=\; c \, V_p \Bigl( \textstyle\sum_{g=1}^{G} U_g U_g^\top \Bigr) V_p \;+\; B_2,
\qquad c = \frac{G}{G-1}, \qquad B_2 = V_p - V_e,
$$
where $c$ is the HC1 finite-sample scaling, $B_2$ is the Marra--Wood Bayesian
term (set $B_2 = 0$ to obtain the purely frequentist sandwich), and the cluster
score contributions $U_g$ differ across estimators. For a single-predictor
exponential family the per-observation score is
$s_i = \bigl[\mu'(\eta_i)(y_i - \mu_i) / (\phi\, V(\mu_i))\bigr] x_i^\top$, with
$x_i$ the $i$th row of $X$, $\phi$ the dispersion, and $V(\cdot)$ the variance
function; for the Gaussian location--scale family (`gaulss`) the location and
scale scores are computed analytically and mapped to coefficient space. The
estimators are:

- **HC** (heteroskedasticity-consistent): each observation is its own cluster,
  $U_i = s_i$, giving the observation-level White/Huber sandwich [@White1980].
  It corrects for heteroskedasticity but, treating grid points as independent,
  *not* for within-curve dependence.
- **Cluster** (CR1): clusters are curves, $U_g = \sum_{i \in g} s_i$
  [@LiangZeger1986; @ZeileisKoellGraham2020]. Summing scores within a curve
  before squaring captures arbitrary within-curve heteroskedasticity and
  autocorrelation, requiring only that distinct curves be independent.
- **CL2** (CR2): a Bell--McCaffrey leverage adjustment of the cluster sandwich
  [@BellMcCaffrey2002; @PustejovskyTipton2018]. With whitened working design
  rows $X_g$ and working residuals $z_g$ for curve $g$, the cluster hat block is
  $H_{gg} = X_g V_p X_g^\top$ and the adjusted contribution is
  $U_g = X_g^\top (I - H_{gg})^{-1/2} z_g$, where $(\cdot)^{-1/2}$ is the
  symmetric inverse square root. The inflation compensates for the downward bias
  of CR1 when the number of clusters $G$ is small.

A subtlety specific to the penalized setting motivates our own implementation
rather than an off-the-shelf one. Generic cluster-robust tools such as
`sandwich::vcovCL()` form the bread from the *unpenalized* cross-product
$(X^\top X)^{-1}$ and compute CR2/HC2 leverages from the unpenalized hat matrix
$X(X^\top X)^{-1} X^\top$. For a penalized fit the relevant operator is instead
the penalized smoother $X V_p X^\top W$, so we use $V_p$ consistently in both the
bread and the cluster leverage $H_{gg} = X_g V_p X_g^\top$. Supplying penalized
fits to `vcovCL()` ignores this and returns invalid covariances, which is why
the estimators above are implemented directly in `refund`.

### Performance measures {#sec-measures}

<!-- TODO: Define coverage, z-scores (z_mean, z_sd), diagonal variance ratio,
     and MC SE computation. Explain that coverage is pointwise over the
     evaluation grid. -->

For each simulation replicate and term type, we compute:

- **Coverage**: the fraction of evaluation grid points where the pointwise CI
  contains the true value.
- **$z$-score diagnostics**: $z_{it} = (\hat\beta_{it} - \beta_{it}^*) /
  \text{SE}_{it}$, summarized as $\bar{z}$ (mean, target 0) and $\text{SD}(z)$
  (target 1). Values of $\text{SD}(z) > 1$ indicate anti-conservative SEs.
- **Diagonal variance ratio**: $\text{mean}(\text{SE}^2) /
  \text{Var}_{\text{MC}}(\hat\beta - \beta^*)$, targeting 1.
  Values below 1 indicate systematic SE underestimation.

Monte Carlo standard errors for coverage are computed as
$\text{SD}(\text{coverage}) / \sqrt{R}$, where each coverage observation is
already a proportion over grid points.


## Study 1: Non-Gaussian Families {#sec-study1}

```{r}
#| label: s1-load
s1 <- load_study1()
```

### Design {#sec-s1-design}

<!-- TODO: Describe the 12 DGP cells, 150 reps, and 4 methods in detail.
     Poisson (log link, additive eps, collapsible) and Binomial (logit,
     Gaussian copula, non-collapsible). -->

Study 1 evaluates CI coverage for `pffr()` under Poisson (log link) and
Binomial (logit link) families. The model includes a function-on-function term
`ff(X1)` and a varying-coefficient linear term `z_lin`.

The factorial design crosses 2 families $\times$ 3 correlation structures (IID,
AR1(0.9), Fourier$^+$(0.3)) $\times$ 2 sample sizes ($n = 200, 400$) = 12 DGP
cells, each with 150 replications. Four CI methods are compared: default, HC,
cluster, and CL2.

Key design decisions:

- **Poisson DGP**: additive $\varepsilon$ on the linear predictor; log link is
  collapsible, so conditional $\beta$ = marginal $\beta$.
- **Binomial DGP**: Gaussian copula preserves marginal $\beta$ (logit is *not*
  collapsible).
- Grid: $30 \times 40$ ($s \times t$). Basis: $k = (12, 12)$ for ff,
  $k = 12$ for linear.

### Results {#sec-s1-results}

```{r}
#| label: s1-summary
s1_summary <- s1 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(family_f, corr_f, n_f, method, term_type) |>
  summarize(
    coverage = mean(coverage, na.rm = TRUE),
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    z_sd = mean(z_sd, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )
```

@fig-s1-heatmap shows coverage across all 12 DGP cells. Under IID errors,
default CIs provide adequate coverage for Binomial but undercover for Poisson
due to unmodeled overdispersion. Under AR1 and Fourier$^+$ correlation, default
coverage collapses across both families, while the cluster sandwich maintains
coverage near 86--89%.

```{r}
#| label: fig-s1-heatmap
#| fig-cap: "Study 1 coverage heatmap. White = 90% nominal. Red = undercoverage."
#| fig-width: 7
#| fig-height: 5
s1_summary |>
  ggplot(aes(
    x = method,
    y = interaction(corr_f, paste0("n=", n_f)),
    fill = coverage
  )) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", coverage * 100)), size = 2.5) +
  scale_fill_gradient2(
    low = "#d73027", mid = "white", high = "#4575b4",
    midpoint = NOMINAL_COVERAGE, limits = c(0.4, 1),
    labels = scales::percent
  ) +
  facet_grid(family_f ~ term_type) +
  labs(x = "Method", y = "Correlation \u00d7 n", fill = "Coverage") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 7))
```

@fig-s1-improvement quantifies the coverage gain from switching to cluster-robust
CIs. The improvement is largest under correlated errors (up to 40 percentage
points for Poisson under AR1) and negligible under IID for Binomial.

```{r}
#| label: fig-s1-improvement
#| fig-cap: "Coverage improvement (cluster $-$ default). Positive values indicate cluster-robust CIs help."
#| fig-width: 7
#| fig-height: 4.5
s1_paired <- s1_summary |>
  dplyr::filter(method %in% c("default", "cluster")) |>
  select(family_f, corr_f, n_f, method, term_type, coverage) |>
  pivot_wider(names_from = method, values_from = coverage) |>
  mutate(diff = cluster - default)

ggplot(
  s1_paired,
  aes(
    x = interaction(corr_f, paste0("n=", n_f)),
    y = diff, fill = family_f
  )
) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(
    yintercept = c(-PRACTICAL_THRESHOLD, PRACTICAL_THRESHOLD),
    linetype = 3, color = "gray50"
  ) +
  facet_wrap(~term_type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Correlation \u00d7 n", y = "Coverage Difference",
    fill = "Family"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

@fig-s1-zsd shows SE calibration via the $z$-score standard deviation. Under
IID, cluster $\text{SD}(z)$ is near 1 for both families. Under correlation,
default and HC show $\text{SD}(z) \gg 1$, confirming that their SEs are too
small.

```{r}
#| label: fig-s1-zsd
#| fig-cap: "SE calibration ($\\text{SD}(z)$, target = 1). Values above 1 indicate anti-conservative SEs."
#| fig-width: 7
#| fig-height: 5
s1_summary |>
  ggplot(aes(x = method, y = z_sd, color = corr_f, shape = term_type)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_hline(yintercept = c(0.85, 1.15), linetype = 3, color = "gray50") +
  facet_grid(family_f ~ n_f) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Method", y = "SD(z)",
    color = "Correlation", shape = "Term"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

### CL2 at large G {#sec-s1-cl2}

<!-- TODO: Discuss why CL2 correction is negligible at G=200-400 (per-cluster
     leverage tiny), contrasting with Study 2 where G=20-80. -->

At $G = 200$--$400$ (Study 1), the per-cluster leverage is small
($\text{trace}(H_{gg}) \approx 0.22$), so the CL2 bias correction produces
negligible SE inflation. @fig-s1-cl2 confirms that CL2 and cluster coverage are
nearly indistinguishable, consistent with theoretical expectations.

```{r}
#| label: fig-s1-cl2
#| fig-cap: "CL2 vs.~cluster coverage at $G = 200$--$400$. The CL2 correction is negligible at these cluster counts."
#| fig-width: 7
#| fig-height: 5
if ("cl2" %in% levels(s1$method)) {
  s1_cl2_comparison <- s1_summary |>
    dplyr::filter(method %in% c("cluster", "cl2")) |>
    select(family_f, corr_f, n_f, method, term_type, coverage, mc_se)

  ggplot(
    s1_cl2_comparison,
    aes(x = corr_f, y = coverage, color = method, group = method)
  ) +
    geom_point(size = 3, position = position_dodge(width = 0.4)) +
    geom_line(linewidth = 0.6, position = position_dodge(width = 0.4)) +
    geom_errorbar(
      aes(
        ymin = coverage - 1.96 * mc_se,
        ymax = coverage + 1.96 * mc_se
      ),
      width = 0.2, linewidth = 0.5,
      position = position_dodge(width = 0.4)
    ) +
    geom_hline(
      yintercept = NOMINAL_COVERAGE,
      linetype = "dashed", color = "red"
    ) +
    facet_grid(family_f ~ term_type + n_f) +
    scale_y_continuous(labels = scales::percent, limits = c(0.7, 1)) +
    scale_color_manual(values = COLORS_S1[c("cluster", "cl2")]) +
    labs(
      x = "Correlation", y = "Coverage", color = "Method"
    ) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
    )
}
```

### Discussion {#sec-s1-discussion}

<!-- TODO: Summarize Study 1 findings. Key messages:
     - Cluster sandwich essential under within-curve correlation
     - HC corrects heteroskedasticity only, not dependence
     - CL2 negligible at large G
     - Poisson overdispersion requires sandwich even under IID
     - Cluster coverage gap of ~84-89% (not reaching nominal) likely due to
       penalization/smoothing bias, not finite-sample sandwich bias -->


## Study 2: Grid Refinement {#sec-study2}

```{r}
#| label: s2-load
s2 <- load_study2()
s2_grid_levels <- attr(s2, "grid_levels")
s2_grid_labels <- attr(s2, "grid_labels")
s2_cov <- load_study2_cov(s2_grid_levels, s2_grid_labels)
```

### Design {#sec-s2-design}

<!-- TODO: Describe the full factorial design: 3 corr x 3 n x 2 SNR x 3 nygrid
     = 54 cells, 50 reps each, 4 methods. Gaussian family, wiggliness = 5. -->

Study 2 investigates how response grid resolution affects CI quality. The design
is fully factorial: 3 correlation structures (IID, AR1(0.9), Fourier$^+$(0.3))
$\times$ 3 sample sizes ($n = 20, 40, 80$) $\times$ 2 SNR levels (3, 25)
$\times$ 3 response grids ($n_y = 40, 80, 120$ with $n_x$ fixed at 60) = 54
cells, each with 50 replications. Four methods are compared: default, cluster,
HC, and CL2.

Data are generated on the finest grid ($90 \times 120$) and deterministically
subsampled for coarser grids, ensuring paired comparisons.

### Results {#sec-s2-results}

```{r}
#| label: s2-summary
s2_summary_by_n <- s2 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(n_f, snr_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    z_mean_sd = sd(z_mean, na.rm = TRUE),
    z_mean = mean(z_mean, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )

s2_summary <- s2 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(snr_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    z_mean_sd = sd(z_mean, na.rm = TRUE),
    z_mean = mean(z_mean, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )

s2_cov_by_term <- s2_summary_by_n |>
  group_by(n_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sqrt(sum(mc_se^2, na.rm = TRUE)) / n(),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    .groups = "drop"
  )
```

@fig-s2-hero presents the central finding: under within-curve correlation,
default SE calibration deteriorates dramatically as the response grid becomes
finer. Coverage of the `ff` term drops from approximately 62% on the coarse grid
to 43% on the fine grid under AR1(0.9), while the cluster sandwich remains
stable. CL2 consistently outperforms cluster by 5--10 percentage points at
$G = 20$--$80$.

```{r}
#| label: fig-s2-hero
#| fig-cap: "Default SE collapse under correlation on finer grids. Red dashed = 90% nominal. SNR averaged. CL2 outperforms cluster by 5--10pp at $G = 20$--$80$."
#| fig-width: 7
#| fig-height: 7
s2_cov_by_term |>
  dplyr::filter(term_type == "ff") |>
  ggplot(aes(
    x = grid_f, y = coverage,
    color = method, group = method
  )) +
  geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_errorbar(
    aes(
      ymin = coverage - 1.96 * mc_se,
      ymax = coverage + 1.96 * mc_se
    ),
    width = 0.2, linewidth = 0.5,
    position = position_dodge(width = 0.3)
  ) +
  geom_hline(
    yintercept = NOMINAL_COVERAGE,
    linetype = "dashed", color = "red", linewidth = 0.8
  ) +
  facet_grid(corr_f ~ n_f) +
  scale_y_continuous(labels = scales::percent, limits = c(0.2, 1)) +
  scale_color_manual(values = COLORS_S2) +
  labs(
    x = "Response Grid (nygrid)", y = "Coverage", color = "Method"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

@fig-s2-diag-ratio provides the mechanistic explanation: the diagonal variance
ratio (model $\text{SE}^2$ / MC variance) drops sharply with finer grids for
default SEs under correlation, while cluster remains closer to 1.

```{r}
#| label: fig-s2-diag-ratio
#| fig-cap: "Diagonal variance ratio by grid resolution. Target = 1. Under correlation, default ratio drops with finer grids; cluster remains stable."
#| fig-width: 7
#| fig-height: 7
if (!is.null(s2_cov)) {
  s2_cov_agg <- s2_cov |>
    group_by(snr_f, corr_f, grid_f, term_type, method) |>
    summarize(
      diag_ratio = median(diag_ratio_median, na.rm = TRUE),
      diag_ratio_lo = quantile(diag_ratio_median, 0.25, na.rm = TRUE),
      diag_ratio_hi = quantile(diag_ratio_median, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(
    s2_cov_agg |> dplyr::filter(term_type != "intercept"),
    aes(x = grid_f, y = diag_ratio, color = method, group = method)
  ) +
    geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
    geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
    geom_errorbar(
      aes(ymin = diag_ratio_lo, ymax = diag_ratio_hi),
      width = 0.2, linewidth = 0.5,
      position = position_dodge(width = 0.3)
    ) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    facet_grid(snr_f + corr_f ~ term_type) +
    scale_color_manual(values = COLORS_S2) +
    labs(
      x = "Response Grid (nygrid)", y = "Diagonal Ratio", color = "Method"
    ) +
    coord_cartesian(ylim = c(0, 2.5)) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
    )
}
```

### Intercept discretization artifact {#sec-s2-intercept}

<!-- TODO: Explain the paradox: coarse grids show ~60-68% intercept coverage
     despite z_mean ~ 0 and z_sd ~ 1. Spatially structured bias alternates in
     sign, canceling in the average. Resolution: threshold around 60x80. -->

On the coarsest grid ($n_y = 40$), intercept coverage drops sharply despite
near-zero mean $z$-scores, an apparent paradox. @fig-s2-intercept reveals the
mechanism: the between-replicate standard deviation of $\bar{z}$ is large on
coarse grids, indicating systematic rep-to-rep intercept shifts that cancel in
the average but cause undercoverage per replicate. The artifact resolves at
$n_y \geq 80$.

```{r}
#| label: fig-s2-intercept
#| fig-cap: "Intercept discretization artifact: between-rep SD of $\\bar{z}$ is large on coarse grids, indicating systematic intercept offsets that cancel on average."
#| fig-width: 7
#| fig-height: 4.5
s2_intercept <- s2_summary |>
  dplyr::filter(term_type == "intercept")

ggplot(
  s2_intercept,
  aes(x = grid_f, y = z_mean_sd, color = method, group = method)
) +
  geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_grid(snr_f ~ corr_f) +
  scale_color_manual(values = COLORS_S2) +
  labs(
    x = "Response Grid (nygrid)",
    y = "SD(z-mean) across reps",
    color = "Method"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

### CL2 vs. cluster at small G {#sec-s2-cl2}

<!-- TODO: Discuss how CL2 outperforms cluster by 5-10pp at G=20-80.
     The CR2 leverage correction produces substantial SE inflation that
     compensates for finite-sample sandwich bias. -->

@fig-s2-cl2-gain shows the incremental coverage gain of CL2 over cluster. Under
correlation, CL2 consistently adds 5--10 percentage points, with the largest
gains at small $n$ (few clusters). Under IID, CL2 also improves over cluster
because the cluster sandwich undercovers due to using $z = 1.96$ critical values
instead of the $t_{G-1}$ distribution at small $G$.

```{r}
#| label: fig-s2-cl2-gain
#| fig-cap: "Coverage gain of CL2 over cluster by grid level. Positive = CL2 improves. Averaged over $n$."
#| fig-width: 7
#| fig-height: 4.5
if (sum(as.character(s2$method) == "cl2", na.rm = TRUE) > 0) {
  s2_paired_cl2 <- build_s2_paired_diff_summary(
    s2_data = s2, num_method = "cl2", den_method = "cluster"
  )

  if (nrow(s2_paired_cl2) > 0) {
    ggplot(s2_paired_cl2, aes(x = grid_f, y = mean_diff, fill = corr_f)) +
      geom_col(position = position_dodge(width = 0.7), width = 0.6) +
      geom_errorbar(
        aes(
          ymin = mean_diff - 1.96 * se_diff,
          ymax = mean_diff + 1.96 * se_diff
        ),
        position = position_dodge(width = 0.7), width = 0.2
      ) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      facet_grid(snr_f ~ term_type) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
      scale_fill_brewer(palette = "Set1") +
      labs(
        x = "Response Grid (nygrid)",
        y = "Coverage Difference",
        fill = "Correlation"
      ) +
      theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
      )
  }
}
```

### Computational cost {#sec-s2-cost}

<!-- TODO: Report fitting times for the four methods across grid sizes.
     Cluster sandwich adds ~X% overhead; CL2 adds more due to leverage
     computation scaling with nygrid. -->

### Discussion {#sec-s2-discussion}

<!-- TODO: Summarize Study 2 findings and connect to practical guidance. -->


## Discussion {#sec-discussion}

<!-- TODO: Synthesize findings across both studies. Key themes:

1. Default SE collapse mechanism: more grid points per curve amplify
   within-curve correlation that default SEs ignore. Diagonal variance ratio
   drops from ~0.3 (coarse) to ~0.1 (fine) for default.

2. The 84-89% cluster coverage gap: even with correct sandwich SEs, cluster
   coverage does not reach 90% nominal. This is NOT finite-sample sandwich
   bias (CL2 doesn't fully close the gap). Likely causes: penalization/
   smoothing bias, penalty-induced distributional effects, grid discretization.

3. CL2 G-dependence: negligible at G=200-400 (Study 1), substantial 5-10pp
   improvement at G=20-80 (Study 2).

4. Practical recommendations: always use cluster-robust sandwich when
   within-curve dependence is possible; prefer CL2 when G < 100;
   default CIs are safe only under IID errors; finer grids require sandwich
   more urgently. -->


## Conclusions {#sec-conclusions}

<!-- TODO: Brief actionable summary of practical recommendations for pffr users. -->


## References {.unnumbered}

::: {#refs}
:::


\appendix

## Computational Environment {#sec-environment}

```{r}
#| label: session-info
sessionInfo()
```
