---
title: "Confidence Intervals for Function-on-Function Regression"
author:
  - name: Fabian Scheipl
    affiliations:
      - name: LMU München
        department: Department of Statistics
        city: Munich
        country: Germany
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
bibliography: pffr-ci-refs.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("_pffr-ci-helpers.R")
```

## Introduction {#sec-intro}

Functional data --- repeated measurements that can be viewed as noisy
realisations of underlying smooth curves --- are now routine in biomedical,
environmental, and engineering applications [@RamsaySilverman2005; @Morris2015].
Function-on-function regression, in which both the response and one or more
predictors are functions, is a central tool in this setting
[@RamsayDalzell1991; @IvanescuStaicuScheiplGreven2015; @MullerYao2008;
@KimStaicuMaityCarrollRuppert2018; @SunDuWangMa2018;
@BrockhausScheiplHothornGreven2015]. The
`pffr()` function in the `refund` package [@ScheiplStaicuGreven2015;
@Rpkg_refund] fits such models by recasting them as large penalized generalized
additive models within the `mgcv` framework [@Wood2017; @WoodPyaSafken2016], a
strategy that unifies a broad family of functional regression models under one
estimation engine [@GrevenScheipl2017].

The point-estimation properties of these models are well understood, but their
*inferential* properties have received far less attention than for their
scalar-on-function counterparts [@ReissGoldsmithShangOgden2017]. By default,
`pffr()` reports pointwise confidence intervals (CIs) built from the Bayesian
posterior covariance matrix of the penalized fit. These intervals enjoy a strong
justification --- they achieve close-to-nominal *across-the-function* frequentist
coverage [@MarraWood2012; @RuppertWandCarroll2003; @Wood2013] --- but only when
the working error model is correctly specified. In functional data this
assumption is routinely violated: the many evaluation points of a single curve
are strongly dependent, yet the default working model treats them as independent
up to a common variance. When such within-curve dependence is present, the
resulting CIs can undercover substantially.

A natural remedy is to replace the model-based covariance with a robust
(sandwich) estimator. Heteroskedasticity-consistent estimators [@White1980] and
cluster-robust estimators that treat each curve as a cluster
[@LiangZeger1986; @CameronMiller2015; @ZeileisKoellGraham2020] require only that
distinct curves be independent, accommodating arbitrary within-curve
heteroskedasticity and autocorrelation. Because the number of curves (clusters)
$G$ can be modest, finite-sample bias of the cluster sandwich matters; the
Bell--McCaffrey/CR2 leverage adjustment [@BellMcCaffrey2002; @PustejovskyTipton2018]
and a $t_{G-1}$ (rather than normal) reference distribution
[@ImbensKolesar2016] are the standard corrections. Resampling alternatives ---
the cluster/curve bootstrap [@FieldWelsh2007] and the wild cluster bootstrap
[@CameronGelbachMiller2008; @MacKinnonWebb2017] --- provide a further route to
robust intervals.

Robust and resampling-based inference is comparatively well developed for the
*functional mean* and for scalar-on-function and function-on-scalar models,
where a sizeable literature constructs pointwise and *simultaneous* confidence
bands: FPCA-corrected bands [@GoldsmithGrevenCrainiceanu2013], tube- and
kinematic-formula bands [@SunLoader1994; @KrivobokovaKneibClaeskens2010;
@Degras2011], geometric and "fast-and-fair" bands for functional parameters
[@ChoiReimherr2018; @LieblReimherr2023], wild-bootstrap bands for
function-on-scalar coefficients [@ChangLinOgden2017], and interval-wise testing
[@PiniVantini2017]. Inference for the bivariate coefficient *surface* of a
function-on-function model is far less developed, with @DetteTang2024 a recent
exception in the unpenalized RKHS setting. To our knowledge, robust pointwise
inference for *penalized* function-on-function regression --- the workhorse
`pffr()` setting --- had not been systematically studied before this work, and
no implementation had been available in widely used software.

This report takes a first step in that direction. We adapt observation-level
(HC), cluster-robust (CR1), and leverage-adjusted (CR2/CL2) sandwich covariance
estimators to the penalized function-on-function setting, clustering by curve,
and evaluate them through two simulation studies reported under the ADEMP
framework [@MorrisWhiteCrowther2019]. Our contributions are threefold: (i) we
provide penalized-GAM cluster and CR2 sandwich estimators for `pffr()`, together
with the corresponding software (@sec-sandwich); (ii) we document a previously
unreported phenomenon whereby the default standard errors *collapse* on finer
response grids under within-curve correlation, so that refining the measurement
grid --- usually thought benign --- actively degrades coverage (@sec-study2);
and (iii) we give practical guidance on when robust intervals are needed and
which variant to use. Study 1 examines non-Gaussian families (Poisson, Binomial)
under varying correlation structures; Study 2 isolates the interaction between
response-grid resolution and CI quality.


## Methods {#sec-methods}

### The pffr model {#sec-pffr}

We adopt the general framework for functional regression of
@GrevenScheipl2017, which embeds function-on-function regression in a unified
class of penalized generalized additive models. For curve (unit) $i =
1,\ldots,G$ we observe a response $Y_i(t)$ over a functional domain $t \in
\mathcal{T}$, evaluated at grid points $t_{id}$, $d = 1,\ldots,D_i$. Conditional
on covariates, $Y_i(t)$ follows an exponential-family distribution whose mean is
linked to an additive predictor $\eta_i(t)$ through a known link function $g$,
$$
g\bigl(\mathbb{E}[Y_i(t) \mid \mathbf{x}_i]\bigr) = \eta_i(t)
  = \sum_{j=1}^{J} h_j(\mathbf{x}_i)(t),
$$
where each partial effect $h_j(\mathbf{x}_i)(t)$ is a smooth function of $t$ and
of a (possibly empty) subset of the scalar and functional covariates collected
in $\mathbf{x}_i$. Special cases of the partial effects $h_j$ include the
functional intercept $\beta_0(t)$ (depending on no covariate); the
function-on-function term $\int_{\mathcal{S}} X_i(s)\,\beta(s,t)\,ds$ with
bivariate coefficient surface $\beta(s,t)$; the function-on-scalar /
varying-coefficient term $z_i\,\beta(t)$ for a scalar covariate $z_i$; the
concurrent effect $X_i(t)\,\beta(t)$; the smooth interaction $f(z_i, t)$; and
*functional random effects* such as subject-specific functional intercepts
$B_a(t)$, which @GrevenScheipl2017 treat as a special class of partial effects
(an indicator basis in the covariate direction with a precision-matrix penalty
in place of a smoothness penalty). Explicit additive residuals appear only in
the Gaussian-identity special case, where the model reduces to
$$
Y_i(t) = \eta_i(t) + \varepsilon_i(t),
\qquad
\eta_i(t) = \beta_0(t) + \int_{\mathcal{S}} X_i(s)\,\beta(s,t)\,ds
  + z_i\,\beta(t) + f(z_i, t) + X_i(t)\,\beta(t).
$$

Crucially, the pffr likelihood *always* assumes that the per-$(i,t)$
contributions are independent conditional on the additive predictor: in the
words of @GrevenScheipl2017, the observed response values are independent "within
and across functions conditional on the additive predictor". The working
likelihood therefore treats the many evaluation points $\{Y_i(t_{id})\}_{d}$ of a
single curve as conditionally independent; any within-curve dependence in the
residual process is *not* modelled, and only explicitly included functional
random effects induce marginal within-curve dependence. This
conditional-independence assumption --- routinely violated by the smooth,
strongly autocorrelated residual processes of real functional data --- is exactly
what the sandwich estimators of @sec-sandwich are designed to correct.

For estimation, all $N = \sum_{i=1}^{G} D_i$ scalar response evaluations are
stacked into a single response vector and the partial effects are represented in
spline bases, so that $g(\mathbb{E}[\mathbf{Y}]) = \mathbf{X}\boldsymbol{\theta}$
with stacked design matrix $\mathbf{X}$ and coefficient vector
$\boldsymbol{\theta}$ collecting the basis coefficients of every term. Effects
that vary over both a covariate and $t$ (such as $\beta(s,t)$ or $f(z_i,t)$) are
built from tensor-product (row-Kronecker) bases, and each term carries a
quadratic penalty $\boldsymbol{\theta}_j^\top \mathbf{P}_j \boldsymbol{\theta}_j$
with marginal smoothing parameters selected by (RE)ML [@Wood2017;
@WoodPyaSafken2016]. The model is fitted as one large penalized GAM via
`mgcv::gam()`, with the usual sum-to-zero and marginal identifiability
constraints on the partial effects [@ScheiplGreven2016].

### Sandwich covariance estimators {#sec-sandwich}

#### The stacked penalized fit and its model-based covariances

After basis expansion (@sec-pffr), `pffr()` stacks all $N = \sum_{i=1}^{G} D_i$
scalar response evaluations (curve $i$ observed at $D_i$ grid points) into a
single penalized GAM with design matrix $\mathbf{X}$ ($N \times p$), coefficient
vector $\hat{\boldsymbol{\theta}}$ collecting the spline coefficients of every
model term, and a block-diagonal smoothing penalty
$\mathbf{S} = \sum_j \lambda_j \mathbf{S}_j$. The fit maximises the penalized
log-likelihood by penalized iteratively reweighted least squares (P-IRLS), which
at convergence defines a diagonal working-weight matrix $\mathbf{W}$ with
entries $w_{id} = \{(d\mu/d\eta)^2 / [\phi\,V(\mu)]\}_{id}$ and a working
response. `mgcv` returns two covariance matrices for
$\hat{\boldsymbol{\theta}}$: the **Bayesian posterior** covariance
$\mathbf{V}_p$ --- formally
$(\mathbf{X}^\top \mathbf{W} \mathbf{X} + \mathbf{S})^{-1}$, though in practice
`mgcv` returns the posterior covariance of the reparameterized coefficients and
this raw inverse need not exist for rank-deficient or constrained penalties ---
and the **frequentist** covariance
$\mathbf{V}_e = \mathbf{V}_p (\mathbf{X}^\top \mathbf{W} \mathbf{X}) \mathbf{V}_p$.
Default `pffr()` CIs use $\mathbf{V}_p$, which delivers good
*across-the-function* frequentist coverage under a correctly specified error
model [@Nychka1988; @MarraWood2012]. The key identity is
$\mathbf{B}_2 \;{:=}\; \mathbf{V}_p - \mathbf{V}_e = \mathbf{V}_p \mathbf{S}
\mathbf{V}_p \succeq 0$ (positive semi-definite because the penalty
$\mathbf{S} \succeq 0$): this penalty-induced (Marra--Wood squared-bias)
component [@MarraWood2012] inflates the Bayesian intervals just enough to achieve
average frequentist coverage. This guarantee, however, presumes independent,
correctly weighted residuals; it says nothing about the within-curve dependence
that the conditional-independence assumption of @sec-pffr ignores, and which is
the norm in functional data.

#### A score (estimating-equation) view {#sec-scores}

All robust estimators replace the model-based *meat* of the covariance by an
empirical outer product of cluster-aggregated scores, while retaining the
*penalized bread* $\mathbf{V}_p$ and the penalty-induced correction
$\mathbf{B}_2$. They share the form
$$
\widehat{\mathbf{V}} \;=\; c \, \mathbf{V}_p
  \Bigl( \textstyle\sum_{g=1}^{G} \mathbf{U}_g \mathbf{U}_g^\top \Bigr)
  \mathbf{V}_p \;+\; \mathbf{B}_2,
\qquad c = \frac{G}{G-1}, \qquad \mathbf{B}_2 = \mathbf{V}_p \mathbf{S} \mathbf{V}_p,
$$ {#eq-sandwich}
where $c = G/(G-1)$ is a simplified cluster correction (analogous to CR1, *not*
the HC1 factor $N/(N-p)$), $\mathbf{B}_2 = \mathbf{V}_p \mathbf{S} \mathbf{V}_p$
is the Marra--Wood penalty-induced bias term (set $\mathbf{B}_2 = \mathbf{0}$ to
obtain the purely frequentist sandwich), and the cluster contributions
$\mathbf{U}_g$ differ across estimators. The building block is the
per-observation score vector $\mathbf{s}_{id}$, the gradient of the
$(i,d)$ log-likelihood contribution with respect to $\boldsymbol{\theta}$,
written here as a length-$p$ row of the score matrix.

*Gaussian/identity case.* With $\mu_{id} = \mathbf{x}_{id}^\top
\boldsymbol{\theta}$ and dispersion $\phi$, the score reduces to
$\mathbf{s}_{id} = \phi^{-1} (y_{id} - \mu_{id})\,\mathbf{x}_{id}$, where
$\mathbf{x}_{id}$ is the corresponding row of $\mathbf{X}$.

*General exponential-family / GLM case.* For a one-parameter exponential family
with link $g$, mean $\mu_{id}$, variance function $V(\cdot)$ and dispersion
$\phi$, the per-observation score is
$$
\mathbf{s}_{id}
  = \frac{(d\mu/d\eta)_{id}\,(y_{id} - \mu_{id})}{\phi\, V(\mu_{id})}\;\mathbf{x}_{id},
$$
i.e. the GLM working residual times the design row. In the implementation the
multiplier $(d\mu/d\eta)(y-\mu)/[\phi V(\mu)]$ is evaluated from the fitted
`mgcv` object via `family$mu.eta`, `family$variance`, and the estimated scale.

*Gaussian location--scale case (`gaulss`).* The `gaulss` family models *both*
the conditional mean and the conditional (log) standard deviation with two
linear predictors, a location predictor $\eta^{(1)}_{id}$ and a scale predictor
$\eta^{(2)}_{id}$, parameterised in `mgcv` through $\tau = 1/\sigma$ with a
`logb` link on $\tau$. Because `gaulss` supplies its own `family$sandwich`
method, the generic GLM score path above does not apply; the two score
components are instead derived analytically from the location--scale
log-likelihood and mapped from natural-parameter space to coefficient space by
the chain rule through `mu.eta`. With residual $r_{id} = y_{id} - \mu_{id}$, the
location and scale score weights are
$$
w^{(1)}_{id} = \tau_{id}^2\, r_{id}\,\bigl(d\mu/d\eta^{(1)}\bigr)_{id},
\qquad
w^{(2)}_{id} = \Bigl(\tfrac{1}{\tau_{id}} - \tau_{id}\, r_{id}^2\Bigr)
  \bigl(d\tau/d\eta^{(2)}\bigr)_{id},
$$
and these weights multiply the location- and scale-block columns of
$\mathbf{X}$, respectively (the two linear predictors index disjoint coefficient
blocks via the `lpi` attribute), to give the full score row $\mathbf{s}_{id}$.
Prior weights, when present, scale both components.

#### The three estimators {#sec-three-estimators}

- **HC** (heteroskedasticity-consistent): each observation is its own cluster,
  $\mathbf{U}_{id} = \mathbf{s}_{id}$, giving the observation-level White/Huber
  sandwich [@White1980] (computed via `mgcv::vcov.gam(sandwich = TRUE)`). It
  corrects for heteroskedasticity but, treating grid points as independent,
  *not* for within-curve dependence.
- **Cluster** (CR1): clusters are curves, so the within-curve scores are summed
  before squaring,
  $\mathbf{U}_g = \sum_{d=1}^{D_g} \mathbf{s}_{gd}$
  [@LiangZeger1986; @ZeileisKoellGraham2020], and the meat is
  $\sum_{g} \mathbf{U}_g \mathbf{U}_g^\top$ with the $c = G/(G-1)$ factor. Summing
  within a curve captures arbitrary within-curve heteroskedasticity and
  autocorrelation, requiring only that distinct curves be independent. (We use
  the simplified CR1 factor rather than the leverage-based HC1 correction.)
- **CL2** (CR2): a Bell--McCaffrey leverage adjustment of the cluster sandwich
  [@BellMcCaffrey2002; @PustejovskyTipton2018], which inflates the contributions
  to compensate for the downward bias of CR1 at small $G$. The score is
  factorised as $\mathbf{s}_{gd} = z_{gd}\,\tilde{\mathbf{x}}_{gd}$ into a scalar
  whitened working residual $z_{gd}$ and an IRLS working-weighted ("whitened")
  design row $\tilde{\mathbf{x}}_{gd}$ (for the GLM case,
  $\tilde{\mathbf{x}}_{gd} = \sqrt{w_{gd}}\,\mathbf{x}_{gd}$ and
  $z_{gd} = \sqrt{w_{gd}}\,(y_{gd} - \mu_{gd})$, with
  $w_{gd} = (d\mu/d\eta)^2_{gd} / [\phi V(\mu_{gd})]$). Stacking the rows of
  curve $g$ into $\tilde{\mathbf{X}}_g$ ($D_g \times p$) and $\mathbf{z}_g$, the
  per-cluster hat block is the $D_g \times D_g$ matrix
  $\mathbf{H}_{gg} = \tilde{\mathbf{X}}_g \mathbf{V}_p \tilde{\mathbf{X}}_g^\top$,
  and the adjusted contribution is
  $$
  \mathbf{U}_g = \tilde{\mathbf{X}}_g^\top
    (\mathbf{I} - \mathbf{H}_{gg})^{-1/2}\,\mathbf{z}_g,
  $$
  where $(\cdot)^{-1/2}$ is the *symmetric* inverse square root computed from the
  eigendecomposition. To keep $(\mathbf{I} - \mathbf{H}_{gg})$ invertible when a
  cluster has high leverage, the eigenvalues of $\mathbf{H}_{gg}$ are capped at
  $0.999$ before the adjustment and a small eigenvalue floor is applied during
  the inverse-square-root step. For `gaulss`, an exact CR2 block is not available
  because the score has two heterogeneous components; instead each observation is
  expanded into two pseudo-observations (one per linear predictor) and the
  whitened rows are formed from a *sign-split* factorization $z = \mathrm{sign}(w)
  \sqrt{|w|}$, $\tilde{\mathbf{x}} = \sqrt{|w|}\,\mathbf{x}$ of the location and
  scale score weights $w^{(1)}, w^{(2)}$. This yields an *approximate* leverage
  adjustment rather than the exact Bell--McCaffrey/CR2 block.

An alternative default baseline is `mgcv`'s `unconditional = TRUE` covariance,
which additionally accounts for smoothing-parameter uncertainty via a
Kass--Steffey correction [@WoodPyaSafken2016]; our study focuses on the standard
$\mathbf{V}_p$ default, but comparing against `unconditional = TRUE` is a natural
direction for future work.

#### Why the penalized bread matters {#sec-penalized-bread}

A subtlety specific to the penalized setting motivates our own implementation
rather than an off-the-shelf one. Generic cluster-robust tools such as
`sandwich::vcovCL()` form the bread from the *unpenalized* cross-product
$(\mathbf{X}^\top \mathbf{X})^{-1}$ and compute CR2/HC2 leverages from the
unpenalized hat matrix
$\mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top$. For a penalized
fit the relevant operator is instead the penalized smoother
$\mathbf{X} \mathbf{V}_p \mathbf{X}^\top \mathbf{W}$, so we use $\mathbf{V}_p$
consistently in both the bread of @eq-sandwich and the cluster leverage
$\mathbf{H}_{gg} = \tilde{\mathbf{X}}_g \mathbf{V}_p \tilde{\mathbf{X}}_g^\top$.
Supplying penalized fits to `vcovCL()` ignores this --- mixing a penalized
$\mathbf{V}_p$-based `bread()` with an unpenalized hat matrix can inflate the
HC2/CR2 leverages by orders of magnitude --- and returns invalid covariances,
which is why the estimators above are implemented directly in `refund`.

#### Reference distribution and band type {#sec-inference-choices}

Two further inference choices interact with the covariance estimator. First, the
pointwise critical value: all intervals reported here use a normal reference
($z = 1.96$), but at small cluster counts $G$ a $t_{G-1}$ reference is the
standard small-sample correction in the cluster-robust literature
[@ImbensKolesar2016; @CameronMiller2015], and we examine its potential role when
interpreting the small-$G$ results of @sec-study2. Second, our intervals are
*pointwise* in $t$ rather than *simultaneous* over the whole coefficient
function or surface; constructing simultaneous bands from a robust covariance is
a natural extension that we do not pursue here. Both choices are evaluated
empirically rather than asserted in advance.

### Performance measures {#sec-measures}

For each simulation replicate and term type, we compute:

- **Coverage**: the fraction of evaluation grid points where the pointwise CI
  contains the true value.
- **$z$-score diagnostics**: $z_{it} = (\hat\beta_{it} - \beta_{it}^*) /
  \text{SE}_{it}$, summarized as $\bar{z}$ (mean, target 0) and $\text{SD}(z)$
  (target 1). Values of $\text{SD}(z) > 1$ indicate anti-conservative SEs.
- **Diagonal variance ratio**: $\text{mean}(\text{SE}^2) /
  \text{Var}_{\text{MC}}(\hat\beta - \beta^*)$, targeting 1.
  Values below 1 indicate systematic SE underestimation.

Monte Carlo standard errors for coverage are computed as
$\text{SD}(\text{coverage}) / \sqrt{R}$, where each coverage observation is
already a proportion over grid points.


## Simulation studies {#sec-results}

We evaluate the four CI methods under two complementary simulation studies,
designed and reported following the ADEMP framework of @MorrisWhiteCrowther2019.

### Aims

Both studies target nominal 90% pointwise coverage of each coefficient term
under conditions that are routine in functional data analysis. Study 1 asks
whether cluster-robust (CR1) and leverage-adjusted (CL2) sandwich CIs maintain
coverage when the response family is non-Gaussian (Poisson, Binomial) and
within-curve errors are correlated, using the default Bayesian and observation-level
HC intervals as foils known to undercover under within-curve correlation.
Study 2 asks how coverage and SE calibration change as the response evaluation
grid is refined, and in particular whether finer grids amplify the failure of
default SEs under within-curve dependence.

### Data-generating mechanisms

Both studies share the same general `pffr()` model structure: curves $i = 1,
\ldots, G$ are observed over a response domain $t \in \mathcal{T}$ and a
predictor domain $s \in \mathcal{S}$; the linear predictor contains at minimum
a functional intercept $\beta_0(t)$, a function-on-function term
$\int X_{1,i}(s)\,\beta(s,t)\,ds$, and a varying-coefficient linear
term $z_{\text{lin},i}\,\beta_{\text{lin}}(t)$. Study 2 additionally includes
a smooth interaction $f(z_{\text{smoo},i}, t)$ and a concurrent term
$X_{\text{conc},i}(t)\,\beta_{\text{conc}}(t)$.

Three within-curve error structures are common to both studies:

- **IID**: $\varepsilon_i(t) \overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2)$ (diagonal covariance).
- **AR1(0.9)**: $\text{Cov}(\varepsilon_i(t_j), \varepsilon_i(t_k)) = \sigma^2 \, 0.9^{|j-k|}$.
- **Fourier$^+$(0.3)**: $\text{Cov}(\varepsilon_i(t_j), \varepsilon_i(t_k)) = \sigma^2 \max(0, \cos(2\pi|t_j - t_k|/0.3))$; non-monotone, always non-negative.

True coefficient functions and surfaces are random draws from tensor B-spline
bases (6 knots per margin), re-scaled so that each term contributes comparable
variation ($k_{\text{truth}} = 6$; fitting basis $k = 12$ to limit smoothing bias).

**Study 1** is fully factorial over:
2 families (Poisson log; Binomial logit) $\times$
3 correlation structures $\times$
2 sample sizes ($G \in \{200, 400\}$)
$= \mathbf{12}$ **DGP cells**, each with $R = 150$ replications ($1{,}800$ total
fits). The predictor grid is fixed at $n_x = 30$; the response grid at $n_y = 40$.
For Poisson, the log link is collapsible, so conditional and marginal $\beta$ coincide.
For Binomial, a Gaussian copula construction preserves the marginal logit
$\beta$ despite the non-collapsibility of the logit link.

**Study 2** is fully factorial over:
3 correlation structures $\times$
3 sample sizes ($G \in \{20, 40, 80\}$) $\times$
2 SNR levels (25 [high], 3 [low])
$= \mathbf{18}$ **DGP cells**, each evaluated at 3 response grid resolutions
($n_y \in \{40, 80, 120\}$), giving $54$ (DGP, grid) combinations with $R = 50$
replications each ($2{,}700$ total fits). The predictor grid is fixed at $n_x = 60$.
Data are generated once on the finest grid ($90 \times 120$) and deterministically
subsampled (evenly spaced indices) to coarser resolutions, ensuring paired
comparisons across grids.

A side-by-side comparison of the two designs:

| Feature | Study 1 | Study 2 |
|---|---|---|
| Response family | Poisson, Binomial | Gaussian |
| Model terms | ff + linear | ff + linear + smooth + concurrent |
| $G$ (curves) | 200, 400 | 20, 40, 80 |
| $n_y$ (response grid) | 40 (fixed) | 40, 80, 120 (varied) |
| $n_x$ (predictor grid) | 30 | 60 |
| SNR varied | No | Yes (25, 3) |
| DGP cells | 12 | 18 × 3 grids = 54 |
| Reps per cell | 150 | 50 |
| Total fits | 1,800 | 2,700 |
| Covariance-quality diagnostics | No | Yes |

: Comparison of Study 1 and Study 2 simulation designs. {#tbl-design-comparison}

### Estimands

In both studies the estimands are the pointwise values of each coefficient
function or surface at the fitted evaluation grid: the bivariate ff surface
$\beta(s,t)$, the linear varying coefficient $\beta_{\text{lin}}(t)$, and the
functional intercept $\beta_0(t)$. Study 2 additionally targets the smooth
$f(z_{\text{smoo}},t)$ and concurrent $\beta_{\text{conc}}(t)$ terms.
The predicted mean response $\mathbb{E}[Y_i(t)]$ (the full $G \times n_y$
fitted surface) is also recorded. Coefficient truths are centred to match
identifiability constraints imposed by the fitting basis (e.g.\ the ff surface
satisfies $\int \beta(s,t)\,ds = 0$; Study 2 terms satisfy sum-to-zero per
evaluation point).

### Methods

All four CI methods are computed from a **single** `pffr()` fit per replicate;
the sandwich corrections are applied post-hoc at the coefficient level:

- **Default**: pointwise CIs from the Bayesian posterior covariance $\mathbf{V}_p$.
- **HC**: observation-level (White/HC) sandwich; corrects for heteroskedasticity
  but not within-curve dependence (@sec-sandwich).
- **Cluster** (CR1): cluster-robust sandwich, treating each curve as one cluster;
  captures arbitrary within-curve heteroskedasticity and autocorrelation.
- **CL2** (CR2): Bell–McCaffrey leverage-adjusted cluster sandwich; reduces
  finite-sample downward bias of CR1 at small $G$ (@sec-sandwich).

All intervals are symmetric and pointwise: $\hat\beta \pm z_{0.95} \cdot \text{SE}$,
with a standard normal critical value $z_{0.95} = 1.96$ ($\alpha = 0.10$) in
both studies.

### Performance measures

Per replicate, per term type, we compute the following performance measures
(following @MorrisWhiteCrowther2019 Section 4):

- **Coverage**: fraction of evaluation grid points where the pointwise CI contains
  the true value; target 0.90. Monte Carlo standard errors are $\text{SD}(\text{coverage})/\sqrt{R}$.
- **$z$-score diagnostics**: $z_{it} = (\hat\beta_{it} - \beta_{it}^*)/\text{SE}_{it}$;
  reported as $\bar z$ (mean; target 0) and $\text{SD}(z)$ (target 1). Values
  $\text{SD}(z) > 1$ indicate anti-conservative SEs.
- **Diagonal variance ratio** (Study 2 only): $\text{mean}(\text{SE}^2) /
  \text{Var}_{\text{MC}}(\hat\beta - \beta^*)$, targeting 1. Values below 1
  indicate systematic SE underestimation; computed by pooling coefficient vectors
  across replications per (DGP, grid, method, term) and summarised as median $\pm$ IQR.
- **Interval width**: mean pointwise CI width, as a secondary measure of
  efficiency.

### Reproducibility

Seeds follow the scheme $\text{seed} = \text{base} + 1000 \cdot \text{DGP\_id} + \text{rep\_id}$,
with base $= 3001$ in Study 1 (DGP 1–12, rep 1–150) and base $= 4001$ in
Study 2 (DGP 1–18, rep 1–50). The same seed is reused across the three
$n_y$ values in Study 2, so that grid comparisons are paired draws from
identical residual processes. Results are saved atomically per replicate (tempfile
+ rename) to support restart without re-running completed replicates.
Parallelisation uses `furrr` with $\leq \min(6, \text{cores}-1)$ workers
(Study 1) or $\leq \min(10, \text{cores}-1)$ (Study 2, one task $=$ one
(DGP, rep) fitting all three grids sequentially).

### Study 1: Non-Gaussian Families {#sec-study1}

```{r}
#| label: s1-load
s1 <- load_study1()
```

#### Design {#sec-s1-design}

Study 1 evaluates CI coverage for `pffr()` under Poisson (log link) and
Binomial (logit link) families, extending the Gaussian-identity setting to
non-Gaussian responses routinely encountered in practice (e.g.\ count and
binary functional outcomes). The model is fitted as `Y ~ ff(X1) + zlin`
with basis dimensions $k = (12, 12)$ for the ff term and $k = 12$ for the
linear term. The factorial design yields 12 DGP cells (2 families $\times$ 3
correlation structures $\times$ 2 sample sizes) with $R = 150$ replications
each; see @tbl-design-comparison for the full factor table.

Key DGP decisions specific to Study 1: for Poisson, an additive Gaussian error
on the linear predictor ensures the log link is collapsible (conditional
$\beta$ = marginal $\beta$, overdispersion $\sigma_\varepsilon = 0.5$); for
Binomial, a Gaussian copula construction preserves the marginal logit $\beta$
despite non-collapsibility of the logit link.

#### Results {#sec-s1-results}

```{r}
#| label: s1-summary
s1_summary <- s1 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(family_f, corr_f, n_f, method, term_type) |>
  summarize(
    coverage = mean(coverage, na.rm = TRUE),
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    z_sd = mean(z_sd, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )
```

@fig-s1-heatmap shows coverage across all 12 DGP cells. Under IID errors,
default CIs provide adequate coverage for Binomial but undercover for Poisson
due to unmodeled overdispersion. Under AR1 and Fourier$^+$ correlation, default
coverage collapses across both families, while the cluster sandwich maintains
coverage near 86--89%.

```{r}
#| label: fig-s1-heatmap
#| fig-cap: "Study 1 coverage heatmap. White = 90% nominal. Red = undercoverage."
#| fig-width: 7
#| fig-height: 5
s1_summary |>
  ggplot(aes(
    x = method,
    y = interaction(corr_f, paste0("n=", n_f)),
    fill = coverage
  )) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", coverage * 100)), size = 2.5) +
  scale_fill_gradient2(
    low = "#d73027", mid = "white", high = "#4575b4",
    midpoint = NOMINAL_COVERAGE, limits = c(0.4, 1),
    labels = scales::percent
  ) +
  facet_grid(family_f ~ term_type) +
  labs(x = "Method", y = "Correlation \u00d7 n", fill = "Coverage") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 7))
```

@fig-s1-improvement quantifies the coverage gain from switching to cluster-robust
CIs. The improvement is largest under correlated errors (up to 40 percentage
points for Poisson under AR1) and negligible under IID for Binomial.

```{r}
#| label: fig-s1-improvement
#| fig-cap: "Coverage improvement (cluster $-$ default). Positive values indicate cluster-robust CIs help."
#| fig-width: 7
#| fig-height: 4.5
s1_paired <- s1_summary |>
  dplyr::filter(method %in% c("default", "cluster")) |>
  select(family_f, corr_f, n_f, method, term_type, coverage) |>
  pivot_wider(names_from = method, values_from = coverage) |>
  mutate(diff = cluster - default)

ggplot(
  s1_paired,
  aes(
    x = interaction(corr_f, paste0("n=", n_f)),
    y = diff, fill = family_f
  )
) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(
    yintercept = c(-PRACTICAL_THRESHOLD, PRACTICAL_THRESHOLD),
    linetype = 3, color = "gray50"
  ) +
  facet_wrap(~term_type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Correlation \u00d7 n", y = "Coverage Difference",
    fill = "Family"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

@fig-s1-zsd shows SE calibration via the $z$-score standard deviation. Under
IID, cluster $\text{SD}(z)$ is near 1 for both families. Under correlation,
default and HC show $\text{SD}(z) \gg 1$, confirming that their SEs are too
small.

```{r}
#| label: fig-s1-zsd
#| fig-cap: "SE calibration ($\\text{SD}(z)$, target = 1). Values above 1 indicate anti-conservative SEs."
#| fig-width: 7
#| fig-height: 5
s1_summary |>
  ggplot(aes(x = method, y = z_sd, color = corr_f, shape = term_type)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_hline(yintercept = c(0.85, 1.15), linetype = 3, color = "gray50") +
  facet_grid(family_f ~ n_f) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Method", y = "SD(z)",
    color = "Correlation", shape = "Term"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

#### CL2 at large G {#sec-s1-cl2}

At $G = 200$--$400$ (Study 1), the per-cluster leverage is small
($\text{trace}(\mathbf{H}_{gg}) \approx 0.22$), so the CL2 bias correction produces
negligible SE inflation. @fig-s1-cl2 confirms that CL2 and cluster coverage are
nearly indistinguishable, consistent with theoretical expectations.

```{r}
#| label: fig-s1-cl2
#| fig-cap: "CL2 vs.~cluster coverage at $G = 200$--$400$. The CL2 correction is negligible at these cluster counts."
#| fig-width: 7
#| fig-height: 5
if ("cl2" %in% levels(s1$method)) {
  s1_cl2_comparison <- s1_summary |>
    dplyr::filter(method %in% c("cluster", "cl2")) |>
    select(family_f, corr_f, n_f, method, term_type, coverage, mc_se)

  ggplot(
    s1_cl2_comparison,
    aes(x = corr_f, y = coverage, color = method, group = method)
  ) +
    geom_point(size = 3, position = position_dodge(width = 0.4)) +
    geom_line(linewidth = 0.6, position = position_dodge(width = 0.4)) +
    geom_errorbar(
      aes(
        ymin = coverage - 1.96 * mc_se,
        ymax = coverage + 1.96 * mc_se
      ),
      width = 0.2, linewidth = 0.5,
      position = position_dodge(width = 0.4)
    ) +
    geom_hline(
      yintercept = NOMINAL_COVERAGE,
      linetype = "dashed", color = "red"
    ) +
    facet_grid(family_f ~ term_type + n_f) +
    scale_y_continuous(labels = scales::percent, limits = c(0.7, 1)) +
    scale_color_manual(values = COLORS_S1[c("cluster", "cl2")]) +
    labs(
      x = "Correlation", y = "Coverage", color = "Method"
    ) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
    )
}
```

#### Discussion {#sec-s1-discussion}

<!-- DRAFT: narrative below uses the established findings recorded in
     PAPER-PLAN.md (Part A4); verify every quoted figure against the
     regenerated Study 1 summaries before submission. -->

Study 1 delivers four messages. First, the cluster-robust sandwich is essential
once within-curve dependence is present: under AR1(0.9) and Fourier$^+$(0.3)
the default intervals collapse for both families, whereas the cluster estimator
holds coverage near approximately 86--89% (@fig-s1-heatmap, @fig-s1-improvement).
Second, the observation-level HC estimator is *not* an adequate substitute ---
because it still treats the grid points within a curve as independent, its
$\text{SD}(z)$ remains well above one under correlation (@fig-s1-zsd), confirming
that the correction must operate at the curve (cluster) level. Third, the CL2
leverage adjustment is negligible at the cluster counts of Study 1
($G = 200$--$400$): the per-cluster leverage is tiny, so CL2 and CR1 coverage
are essentially indistinguishable (@fig-s1-cl2). Fourth, a robust covariance is
sometimes needed even under independent errors: the Poisson default undercovers
under IID because the working model omits overdispersion, while the Binomial
default is adequate under IID, where its mean--variance link is correctly
specified.

A residual gap remains: even the cluster intervals do not reach the nominal 90%,
plateauing around approximately 84--89%. The CL2 correction --- which specifically
targets finite-sample sandwich bias --- does not close this gap, and crucially the
gap persists at the large $G$ of Study 1 ($G = 200$--$400$), where both leverage
and $t$-vs-$z$ reference-distribution effects are negligible. The leading
explanation is penalization/smoothing bias of the coefficient estimates themselves:
the sandwich corrects the *variance* of $\hat{\boldsymbol{\theta}}$ but not the *bias* induced by
the smoothing penalty, which no covariance estimator can remove. Other contributors
(smoothing parameter uncertainty, basis effects) cannot be fully excluded, but
penalization bias has the strongest theoretical support. We return to this point
in @sec-discussion.


### Study 2: Grid Refinement {#sec-study2}

```{r}
#| label: s2-load
s2 <- load_study2()
s2_grid_levels <- attr(s2, "grid_levels")
s2_grid_labels <- attr(s2, "grid_labels")
s2_cov <- load_study2_cov(s2_grid_levels, s2_grid_labels)
```

#### Design {#sec-s2-design}

Study 2 investigates how response grid resolution affects CI quality for the
full `pffr()` model `Y ~ ff(X1) + zlin + s(zsmoo) + Xconc` under Gaussian
errors. The design is summarised in @tbl-design-comparison. The key feature
specific to this study is the paired grid structure: data are generated once
at the finest resolution ($n_x = 90$, $n_y = 120$) and deterministically
subsampled to $n_y \in \{40, 80, 120\}$ (with $n_x$ fixed at 60 in the main
study), ensuring that grid comparisons are paired draws from the same
residual process. The two SNR levels (25 = high, 3 = low) gauge sensitivity
to signal strength; all four CI methods are compared in each cell.

#### Results {#sec-s2-results}

```{r}
#| label: s2-summary
s2_summary_by_n <- s2 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(n_f, snr_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    z_mean_sd = sd(z_mean, na.rm = TRUE),
    z_mean = mean(z_mean, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )

s2_summary <- s2 |>
  dplyr::filter(!is.na(coverage)) |>
  group_by(snr_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sd(coverage, na.rm = TRUE) / sqrt(n()),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    z_mean_sd = sd(z_mean, na.rm = TRUE),
    z_mean = mean(z_mean, na.rm = TRUE),
    width = mean(mean_width, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )

s2_cov_by_term <- s2_summary_by_n |>
  group_by(n_f, corr_f, grid_f, method, term_type) |>
  summarize(
    mc_se = sqrt(sum(mc_se^2, na.rm = TRUE)) / n(),
    coverage = mean(coverage, na.rm = TRUE),
    z_sd = mean(z_sd, na.rm = TRUE),
    .groups = "drop"
  )
```

@fig-s2-hero presents the central finding: under within-curve correlation,
default SE calibration deteriorates dramatically as the response grid becomes
finer. Coverage of the `ff` term drops from approximately 60--65% on the coarse
grid to around 40--45% on the fine grid under AR1(0.9), while the cluster
sandwich remains stable. CL2 consistently outperforms cluster by approximately
5--10 percentage points at $G = 20$--$80$.

```{r}
#| label: fig-s2-hero
#| fig-cap: "Default SE collapse under correlation on finer grids. Red dashed = 90% nominal. SNR averaged. CL2 approximately outperforms cluster by 5--10pp at $G = 20$--$80$."
#| fig-width: 7
#| fig-height: 7
s2_cov_by_term |>
  dplyr::filter(term_type == "ff") |>
  ggplot(aes(
    x = grid_f, y = coverage,
    color = method, group = method
  )) +
  geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_errorbar(
    aes(
      ymin = coverage - 1.96 * mc_se,
      ymax = coverage + 1.96 * mc_se
    ),
    width = 0.2, linewidth = 0.5,
    position = position_dodge(width = 0.3)
  ) +
  geom_hline(
    yintercept = NOMINAL_COVERAGE,
    linetype = "dashed", color = "red", linewidth = 0.8
  ) +
  facet_grid(corr_f ~ n_f) +
  scale_y_continuous(labels = scales::percent, limits = c(0.2, 1)) +
  scale_color_manual(values = COLORS_S2) +
  labs(
    x = "Response Grid (nygrid)", y = "Coverage", color = "Method"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

@fig-s2-diag-ratio provides the mechanistic explanation: the diagonal variance
ratio (model $\text{SE}^2$ / MC variance) drops sharply with finer grids for
default SEs under correlation, while cluster remains closer to 1.

```{r}
#| label: fig-s2-diag-ratio
#| fig-cap: "Diagonal variance ratio by grid resolution. Target = 1. Under correlation, default ratio drops with finer grids; cluster remains stable."
#| fig-width: 7
#| fig-height: 7
if (!is.null(s2_cov)) {
  s2_cov_agg <- s2_cov |>
    group_by(snr_f, corr_f, grid_f, term_type, method) |>
    summarize(
      diag_ratio = median(diag_ratio_median, na.rm = TRUE),
      diag_ratio_lo = quantile(diag_ratio_median, 0.25, na.rm = TRUE),
      diag_ratio_hi = quantile(diag_ratio_median, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(
    s2_cov_agg |> dplyr::filter(term_type != "intercept"),
    aes(x = grid_f, y = diag_ratio, color = method, group = method)
  ) +
    geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
    geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
    geom_errorbar(
      aes(ymin = diag_ratio_lo, ymax = diag_ratio_hi),
      width = 0.2, linewidth = 0.5,
      position = position_dodge(width = 0.3)
    ) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    facet_grid(snr_f + corr_f ~ term_type) +
    scale_color_manual(values = COLORS_S2) +
    labs(
      x = "Response Grid (nygrid)", y = "Diagonal Ratio", color = "Method"
    ) +
    coord_cartesian(ylim = c(0, 2.5)) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
    )
}
```

#### Intercept discretization artifact {#sec-s2-intercept}

On the coarsest grid ($n_y = 40$), intercept coverage drops sharply despite
near-zero mean $z$-scores, an apparent paradox. @fig-s2-intercept reveals the
mechanism: the between-replicate standard deviation of $\bar{z}$ is large on
coarse grids, indicating systematic rep-to-rep intercept shifts that cancel in
the average but cause undercoverage per replicate. The artifact resolves at
$n_y \geq 80$.

```{r}
#| label: fig-s2-intercept
#| fig-cap: "Intercept discretization artifact: between-rep SD of $\\bar{z}$ is large on coarse grids, indicating systematic intercept offsets that cancel on average."
#| fig-width: 7
#| fig-height: 4.5
s2_intercept <- s2_summary |>
  dplyr::filter(term_type == "intercept")

ggplot(
  s2_intercept,
  aes(x = grid_f, y = z_mean_sd, color = method, group = method)
) +
  geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_grid(snr_f ~ corr_f) +
  scale_color_manual(values = COLORS_S2) +
  labs(
    x = "Response Grid (nygrid)",
    y = "SD(z-mean) across reps",
    color = "Method"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

#### CL2 vs. cluster at small G {#sec-s2-cl2}

@fig-s2-cl2-gain shows the incremental coverage gain of CL2 over cluster. Under
correlation, CL2 consistently adds approximately 5--10 percentage points, with
the largest gains at small $n$ (few clusters). Under IID, CL2 also improves over
cluster, possibly because the cluster sandwich undercovers due to using $z = 1.96$
critical values rather than a $t_{G-1}$ distribution at small $G$ (a conjecture
to be confirmed by a direct reference-distribution experiment).

```{r}
#| label: fig-s2-cl2-gain
#| fig-cap: "Coverage gain of CL2 over cluster by grid level. Positive = CL2 improves. Averaged over $n$."
#| fig-width: 7
#| fig-height: 4.5
if (sum(as.character(s2$method) == "cl2", na.rm = TRUE) > 0) {
  s2_paired_cl2 <- build_s2_paired_diff_summary(
    s2_data = s2, num_method = "cl2", den_method = "cluster"
  )

  if (nrow(s2_paired_cl2) > 0) {
    ggplot(s2_paired_cl2, aes(x = grid_f, y = mean_diff, fill = corr_f)) +
      geom_col(position = position_dodge(width = 0.7), width = 0.6) +
      geom_errorbar(
        aes(
          ymin = mean_diff - 1.96 * se_diff,
          ymax = mean_diff + 1.96 * se_diff
        ),
        position = position_dodge(width = 0.7), width = 0.2
      ) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      facet_grid(snr_f ~ term_type) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
      scale_fill_brewer(palette = "Set1") +
      labs(
        x = "Response Grid (nygrid)",
        y = "Coverage Difference",
        fill = "Correlation"
      ) +
      theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
      )
  }
}
```

#### Computational cost {#sec-s2-cost}

The three robust estimators differ markedly in cost, and the difference is
structural rather than incidental. The CR1 cluster sandwich adds only a single
post-fit pass over the data: per-observation scores are aggregated to
cluster totals and combined into a $p \times p$ meat matrix, so its overhead
relative to the default fit is small and essentially independent of the response
grid. The CL2 estimator is more expensive because the leverage adjustment is
computed *per cluster*: for each curve $g$ it forms the $n_g \times n_g$ hat
block $\mathbf{H}_{gg} = \tilde{\mathbf{X}}_g \mathbf{V}_p \tilde{\mathbf{X}}_g^\top$ and its symmetric inverse square root via an
eigendecomposition, where $n_g$ is the number of evaluation points on that
curve. This $O(n_g^3)$ per-cluster step makes CL2 cost grow sharply with the
response-grid resolution $n_y$, exactly the axis Study 2 varies.

<!-- TODO(data): insert the measured fitting times (default / HC / cluster / CL2
     across n_y = 40, 80, 120) from study2-cl2-timing/ once the result
     directories are available in the build environment; report cluster overhead
     as a percentage of the base fit and the CL2 scaling in n_y. -->

#### Discussion {#sec-s2-discussion}

<!-- DRAFT: narrative below uses the established findings recorded in
     PAPER-PLAN.md (Part A4); verify every quoted figure against the
     regenerated Study 2 summaries before submission. -->

<!-- DRAFT: narrative below uses the established findings recorded in
     PAPER-PLAN.md (Part A4); verify every quoted figure against the
     regenerated Study 2 summaries before submission. -->
Study 2 isolates a phenomenon that is easy to miss in practice: refining the
response grid, normally regarded as harmless or even beneficial, actively
*degrades* default coverage under within-curve correlation. For the `ff` term
under AR1(0.9), default coverage falls from approximately 60--65% on the coarse
grid ($n_y = 40$) to around 40--45% on the fine grid ($n_y = 120$), while the
cluster sandwich stays essentially flat near approximately 86--87% (@fig-s2-hero).
The mechanism is visible in the diagonal variance ratio (@fig-s2-diag-ratio): as
the grid is refined, each additional, highly correlated evaluation point is
counted by the default formula as fresh information, so the model-based variance
shrinks far faster than the true sampling variance (the diagonal ratio falling
from around 0.3 on coarse grids toward around 0.1 on fine grids under
correlation). The cluster sandwich, which pools the score over the whole curve
before squaring, is largely resistant to this double-counting.

Two secondary findings complete the picture. The intercept "paradox" on coarse
grids --- low coverage despite $\bar z \approx 0$ and $\text{SD}(z) \approx 1$
--- is a discretization artifact: the per-replicate intercept is systematically
offset in a spatially structured way that cancels in the average but not within
each replicate, and it resolves once $n_y \geq 80$ (@fig-s2-intercept). Finally,
at the small cluster counts of Study 2 ($G = 20$--$80$) the CL2 leverage
adjustment matters, adding approximately 5--10 percentage points of coverage over
CR1 (@fig-s2-cl2-gain); part of this gain reflects the leverage inflation proper
and part may reflect CL2 partially compensating for the use of $z = 1.96$ rather
than a $t_{G-1}$ reference at small $G$ (a conjecture to be confirmed by a direct
$t_{G-1}$ experiment) [@ImbensKolesar2016; @CameronMiller2015].


## Discussion {#sec-discussion}

<!-- DRAFT: synthesis below uses the established findings recorded in
     PAPER-PLAN.md (Part A4); verify quoted figures against regenerated
     summaries before submission. -->

<!-- DRAFT: synthesis below uses the established findings recorded in
     PAPER-PLAN.md (Part A4); verify quoted figures against regenerated
     summaries before submission. -->
Taken together, the two studies tell a consistent story about *why* default
`pffr()` intervals fail and *what* fixes them. The failure mode is a mismatch
between the working error model --- independent grid points with a common
variance --- and the strong within-curve dependence of functional data. Because
the default variance formula treats each evaluation point as independent
information, it understates the sampling variance, and the understatement *grows*
with the number of grid points per curve. This is the unifying mechanism behind
both the family-wide collapse of Study 1 and the grid-refinement collapse of
Study 2, where the diagonal variance ratio for default standard errors falls
from around 0.3 on coarse grids toward around 0.1 on fine grids under
correlation. The cluster-robust sandwich removes the mismatch by requiring only
independence *across* curves, and is correspondingly stable across families,
correlation structures, and grid resolutions.

The residual coverage gap --- cluster intervals plateauing around 84--89% rather
than the nominal 90% --- deserves emphasis precisely because it is *not* fixed by
the sandwich machinery. The strongest evidence that this is not a finite-sample
variance-estimation artifact is that the gap persists at large $G$ (Study 1,
$G = 200$--$400$), where both the per-cluster leverage correction and any
$t$-vs-$z$ reference-distribution effect are negligible. The leading explanation
is therefore smoothing/penalization bias in $\hat{\boldsymbol{\theta}}$ itself: a robust
covariance estimator corrects the spread of the sampling distribution but not its
centre, so penalty-induced bias in the coefficient surface leaks directly into
coverage. We do not claim to have ruled out other contributors --- smoothing
parameter uncertainty, basis and discretization effects, or distributional effects
of the penalty at moderate $G$ --- but penalization bias has the strongest
theoretical grounding in this setting. An orthogonal remedy would be to model the
within-curve error covariance explicitly --- for example via functional mixed
models or covariance smoothing [@CederbaumPouplierHooleGreven2016;
@CederbaumScheiplGreven2018] --- rather than to robustify against it; the
sandwich's appeal is that it requires no such model and is agnostic to the form
of the dependence.

The behaviour of CL2 is governed by the number of clusters. At the large $G$ of
Study 1 the leverage adjustment is negligible, whereas at the small $G$ of
Study 2 it contributes approximately 5--10 percentage points, partly by inflating
for leverage and partly (conjecturally) by compensating for the use of $z = 1.96$
rather than a small-sample $t_{G-1}$ reference distribution --- a direct
comparison of $z$- and $t_{G-1}$-based intervals remains to be run. This mirrors
the cluster-robust inference literature in econometrics, where small-$G$
corrections and the choice of reference distribution are central
[@CameronMiller2015; @ImbensKolesar2016], and where bootstrap alternatives ---
the cluster/curve bootstrap [@FieldWelsh2007] and the wild cluster bootstrap
[@CameronGelbachMiller2008; @MacKinnonWebb2017; @MacKinnonWebb2018] --- offer a
further route that we do not pursue here.

Several limitations frame these conclusions. Our intervals are *pointwise*;
inference targeting the whole coefficient surface jointly requires simultaneous
bands, for which a substantial functional-data literature exists
[@SunLoader1994; @KrivobokovaKneibClaeskens2010; @ChoiReimherr2018;
@LieblReimherr2023; @DetteTang2024] and which the cluster-robust covariance could
in principle feed. The persistent penalization-bias gap, the dependence of CL2
on cluster count, and the super-linear computational cost of CL2 in the response
grid are all genuine constraints rather than tuning issues.


## Conclusions {#sec-conclusions}

For practitioners fitting function-on-function regressions with `pffr()`, the
guidance is simple. The default Bayesian intervals are trustworthy only when the
within-curve errors are close to independent and correctly weighted; this is the
exception, not the rule, in functional data. Whenever within-curve dependence is
plausible, use the cluster-robust (`sandwich = "cluster"`) intervals, which were
stable across every family, correlation structure, and grid resolution we
examined. When the number of curves is small (roughly $G < 100$), prefer the
leverage-adjusted `sandwich = "cl2"` variant, which recovers approximately
5--10 percentage points of coverage. Counter-intuitively, a finer response grid makes
robust intervals *more* necessary, not less, because it amplifies the
double-counting that sinks the default standard errors. Finally, users should not
expect even robust intervals to attain exact nominal coverage: a modest residual
shortfall from smoothing bias remains, and is best addressed by attention to the
penalty and basis dimension rather than by the covariance estimator alone.


## References {.unnumbered}

::: {#refs}
:::


\appendix

## Computational Environment {#sec-environment}

```{r}
#| label: session-info
sessionInfo()
```
